Efficient pedigree recording for fast population genetics simulation

### 2018-01-10 forwards_paper.tex
Refocussed the API overview section.

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\fwdpy}{\texttt{fwdpy11}}
\newcommand{\cpp}{\texttt{C++}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\ftprime}{\texttt{ftprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
% mbox to avoid hyphenating
\newcommand{\taddnode}[1]{\mbox{\texttt{addnode}}(#1)}
\newcommand{\taddedge}[1]{\mbox{\texttt{addedge}}(#1)}
\newcommand{\tsimplify}[1]{\mbox{\texttt{simplify}}(#1)}
\newcommand{\tsort}[1]{\mbox{\texttt{sort}}(#1)}
\newcommand{\tnodetable}[1]{\mbox{\texttt{NodeTable}}(#1)}
\newcommand{\tedgetable}[1]{\mbox{\texttt{EdgeTable}}(#1)}
\DeclareMathOperator{\len}{length}
\DeclareMathOperator{\add}{add} % for adding to tables
\newcommand{\lleft}{\mathrm{left}}  % left label
\newcommand{\lright}{\mathrm{right}}
\newcommand{\lid}{\mathrm{id}}

@@

\algstep{W1.}{Initialisation.}{
Set $\Nt \leftarrow \tnodetable()$, $\mathcal{E}
\leftarrow \mbox{\texttt{EdgeTable}}()$, $t \leftarrow T$, and $j \leftarrow 0$.
 For $0 \leq k < N$, set $P_k \leftarrow \taddnode{\Nt, T}$.
}

@@

%%%%%%
\subsection*{Overview of the API}

These tools are implemented through the \msprime{} Python API,
which provides a powerful platform for working with tree topology and mutation data.
The new portions of \msprime{} that we cover here
are dedicated to tree sequence input and output using simple tables of data,
so we refer to this as the `Tables API'.
The four key tables: nodes, edges, sites and mutations, are described above.

The Tables API is primarily designed to facilitate efficient interchange of
data between programs or between different modules of the same program.
Following current best practises
\plr{citations: Apache arrow, etc},
data is stored in a columnar format,
which allows for very efficient interchange of large amounts of numerical data.
In principle, this enables zero-copy semantics,
where a data consumer can read the information directly
from the memory of a producer without incurring the overhead of a copy.
\jda{TODO citation missing}
Our implementation uses the NumPy C API~\citep{walt2011numpy} to efficiently copy
data from Python into the low-level C library used to manipulate tree sequences.
Since tables are encoded as contiguous arrays of numeric data,
this achieves data transfer rates that would be impossible under
any text-based approach while retaining excellent portability.

This also provides basic input and output operations via NumPy's
array interface, which provides a great deal of flexibility and efficiency,
and makes it straightforward to transfer data from sources
such as HDF5 \citep{hdf5}or Dask~\citep{dask}.
For small scale data and debugging purposes, a simple text format is also supported.

% PLR: this was mentioned above, and belongs in the documentation, I think.
% The API also provides functions to sort tables
% to ensure that the records are in the form required to input
% into an \msprime{} tree sequence object. (Simplification may also be required.)

The \msprime{} Tables API is designed to be easily ``plugged'' into any simulation engine.
One simply needs to fill the relevant columns,
and to send them to \msprime{} at regular intervals for simplification.
For example, in the benchmarking simulations above,
we used the \fwdpp{} \cpp{} API \citep{fwdpp}
to fill arrays of simple structures representing the node and edge data.
Doing so requires no modification of the \fwdpp{} code base.
Rather, we simply need to bookkeep parent/offspring labels,
and perform simple processing of the recombination breakpoints from each mating event.
Using \texttt{pybind11} (\url{https://github.com/pybind/pybind11/}),
the node/edge arrays on the \cpp{} side are made visible, without copy, to Python as a multi-column numpy array,
which is processed by the \msprime{} tables API.
Our implementation also uses \texttt{pybind11}
to create a Python package extending \texttt{fwdpy11}
(a Python package based on \fwdpp{}; \url{https://github.com/molpopgen/fwdpy11}),
% (A neat aside, that we may wish to skip here, is that it may be possible to use the tables
% API from a command-line C/\cpp{} application via an embedded Python interpreter.)

@@

\subsection*{The \cpp{} implementation}

\paragraph{Requirements.}
The \cpp{} implementation uses several packages installed via conda,
except for \msprime{}, which must be installed from GitHub.

\begin{enumerate}
    \item \fwdpy{} 0.13.a0
    \item \fwdpp{} 0.5.7
    \item GNU Scientific Library 1.16
    \item pybind11 2.2.1
    \item GCC 4.8.5
    \item \msprime{} 0.5.0 \jda{Changed these but the actual sims were not run
		    with this version...}
\end{enumerate}

\subsection*{The \simupop{} implementation}

We implemented simulations in a Python script, usable as a command-line utility,
taking input parameters for population size $N$ and scaled recombination $\rho$.
As described in 'Simulation benchmarks', we set scaled mutation $\theta=\rho$,
we used simplify interval of 1000 and simulated for $T=10N$ generations in each
simulation, introducing deleterious mutations at a rate of $\rho/100$ per
generation.
Because \simupop{} requires explicit locations for loci, we set the number of
loci $n=\rho / (4Nr)$ and marked
$k={\rm round}(\mu / (100 (\mu + \mu / 100))) = 0$ of these as selected loci.
\jda{unfortunately probably what we did...}
The selected loci were chosen as a random sample of all the loci.

We coupled \simupop{} to \msprime{} using a Python class \texttt{RecombCollector} 1)
tracks time, and 2) collects and parses recombination information provided by
the \texttt{Recombinator} class of \simupop{}, adding them to a \tnodetable{} and
\tedgetable{} upon which simplification is performed every 1000 years (using
\texttt{msprime.sort\_tables()} and \texttt{msprime.simplify\_tables}).
The tables are initialized using the \texttt{TreeSequence} returned from
\texttt{msprime.simulate($2N$, Ne=$N$, recombination\_rate=$r/2$, length=$n$)}.
The \texttt{simuPOP.Population()} is initialized and tagged with individual IDs,
a mapping of these IDs to the \tnodetable IDs is passed to
an instance of \texttt{RecombCollector} and used internally to update the tables.
The \texttt{simuPOP.Population()} is then evolved.
Before mating: selected mutations are created (using
\texttt{simuPOP.SNPMutator}) and multiplicative selection occurs (using
\texttt{simuPOP.PyMlSelector}) only on the selected loci;
the internal time of the \texttt{RecombCollector} instance is updated.
During mating: male and female parents are drawn randomly, recombination data is
output by \simupop{} to the \texttt{RecombCollector} instance.
After mating: nothing happens 999/1000 years, every 1000 years the tables are
simplifed.

\texttt{RecombCollector} uses the Python class \texttt{ARGrecorder} internally
to add data to the tables and perform simplification.
This class contains the \tnodetable and \tedgetable and also the mapping from
(haploid) ids to ids in the tables.
It also provides methods (\texttt{add\_individual} and \texttt{add\_record})
to append data to the tables.
The only functions of
\texttt{RecombCollector} not delegated to this internal class are computing
haploid IDs from \simupop{}'s diploid IDs and parsing recomination data into
data suitible for use with use with \texttt{add\_individual} and
\texttt{add\_record}; these functions are used to stream data into the tables as
it is received by an instance of \texttt{RecombCollector}.

\paragraph{Requirements.}
The simulations as implemented depend on the class \texttt{RecombCollector} provided by
the Python package \ftprime{} 0.0.6rc0,
available at \url{https://www.github.com/ashander/ftprime}.
Its requirements are encoded in an \texttt{environment.yml} file, compatible
with conda, but \msprime{} must be updated to the version listed below via pip
from GitHub \jda{TODO add command?}.

In brief the requirements are:

\begin{enumerate}
    \item \msprime{} 0.5.0
    \item \simupop{} 1.1.8.3
    \item python 3.5
\end{enumerate}

>>>>

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\fwdpy}{\texttt{fwdpy11}}
\newcommand{\cpp}{\texttt{C++}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\ftprime}{\texttt{ftprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
% mbox to avoid hyphenating
\newcommand{\taddnode}[1]{\mbox{\texttt{addnode}}(#1)}
\newcommand{\taddedge}[1]{\mbox{\texttt{addedge}}(#1)}
\newcommand{\tsimplify}[1]{\mbox{\texttt{simplify}}(#1)}
\DeclareMathOperator{\len}{length}
\DeclareMathOperator{\add}{add} % for adding to tables
\newcommand{\lleft}{\mathrm{left}}  % left label
\newcommand{\lright}{\mathrm{right}}
\newcommand{\lid}{\mathrm{id}}

@@

\algstep{W1.}{Initialisation.}{
Set $\Nt \leftarrow \mbox{\texttt{NodeTable}}()$, $\mathcal{E}
\leftarrow \mbox{\texttt{EdgeTable}}()$, $t \leftarrow T$, and $j \leftarrow 0$.
 For $0 \leq k < N$, set $P_k \leftarrow \taddnode{\Nt, T}$.
}

@@

%%%%%%
\subsection*{Implementation overview}

The facilities for working with succinct tree sequences are implemented as part
of the \msprime\ Python API, which provides a powerful platform for processing
tree topology and mutation data. The new portions of \msprime{} that we discuss
here are dedicated to tree sequence input and output using simple tables of
data, so we refer to this as the `Tables API'. The four key tables (nodes,
edges, sites and mutations) are described above.

The Tables API is primarily designed to facilitate efficient interchange of
data between programs or between different modules of the same program. We
adopted a `columnar' design, i.e., where all the values for a
particular column are stored in adjacent memory locations.
There are many advantages to columnar storage---for example, since adjacent
values in memory are from the same column, they tend to compress well,
and suitable encodings can be chosen on a per-column basis~\citep{abadi2006integrating}.
A particular advantage of this approach is that it enables very
efficient copying of data, and in principle zero-copy data access
(where a data consumer reads directly from the memory of a producer) is possible.
Our implementation uses the NumPy C API~\citep{walt2011numpy} to efficiently copy
data from Python into the low-level C library used to manipulate tree sequences.
This architecture allows for data transfer rates of gigabytes per second
(impossible under any text-based approach), while retaining excellent portability.
NumPy's array interface provides a great deal of flexibility and efficiency,
and makes it straightforward to transfer data from sources
such as HDF5 \citep{hdf5}or Dask~\citep{dask}.
For small scale data and debugging purposes, a simple text based format is also supported.

The \msprime\ Python Tables API provides a general purpose toolkit for importing
and processing succinct tree sequences. Interoperation with Python simulators
is then straightforward. For the proof-of-concept implementations discussed
here, we took two different approaches to doing this. In the first method,
based on \simupop{}, the simulation engine invokes callbacks at critical points
of the simulation, and we infer nodes and edges from the information that is
provided. Rows are appended to the tables one-by-one, and the tables are
periodically sorted and simplified to control memory usage. The second method we developed uses
\texttt{pybind11} (\url{https://github.com/pybind/pybind11/}) to inferface
with the \fwdpp{} \cpp{} API \citep{fwdpp}. No modifications were
required to the \fwdpp{} code base; rather, we simply need to bookkeep parent/offspring labels,
and perform simple processing of the recombination breakpoints from each mating
event to generate node and edge data. This information is then periodically copied
to the \msprime\ Tables API, where it is sorted and simplified.

While these methods work well, and indeed result in significant performance
gains over non-ancestry based simulation, they are far from optimal. Because
forwards time simulation is highly CPU intensive, most simulation engines
are written in C or C++ out of necessity. Therefore, the simplest and most
efficient approach would be to call the \msprime\ C library directly as part of
the core simulation logic, in the manner outlined in Algorithm W. If designed
in this way, the simulation engine would not need to keep copies of the node
and edge information and would not need to infer edges;
even greater gains in computational efficiency could then be obtained. To
facilitate this usage, a standalone, embeddable C library encompassing the core
tree sequence functionality is planned, which should make recording and
outputting tree sequences from any C/C++ code straightforward.

% is designed to be easily ``plugged'' into any simulation engine.
% One simply needs to fill the relevant columns,
% and to send them to \msprime{} at regular intervals for simplification.
% For example, in the benchmarking simulations above,
% we used the \fwdpp{} \cpp{} API \citep{fwdpp}
% to fill arrays of simple structures representing the node and edge data.
% Doing so requires no modification of the \fwdpp{} code base.
% Rather, we simply need to bookkeep parent/offspring labels,
% and perform simple processing of the recombination breakpoints from each mating event.
% Using \texttt{pybind11} (\url{https://github.com/pybind/pybind11/}),
% the node/edge arrays on the \cpp{} side are made visible, without copy, to Python as a multi-column numpy array,
% which is processed by the \msprime{} tables API.
% Our implementation also uses \texttt{pybind11}
% to create a Python package extending \texttt{fwdpy11}
% (a Python package based on \fwdpp{}; \url{https://github.com/molpopgen/fwdpy11}),
% % (A neat aside, that we may wish to skip here, is that it may be possible to use the tables
% % API from a command-line C/\cpp{} application via an embedded Python interpreter.)

@@

\krt{Are we doing a methods section?}

\subsection{Requirements for the \cpp{} implementation}

\begin{enumerate}
    \item \fwdpy{} 0.13.a0
    \item \fwdpp{} 0.5.7
    \item GNU Scientific Library 1.16
    \item pybind11 2.2.1
    \item GCC 4.8.5
    \item \msprime{} master branch \jda{which commit hash or short version number
		    from pip freeze or {\tt msprime.\_\_version\_\_}?}
\end{enumerate}

All of the above installed via conda, except for msprime, which must be installed from GitHub.

\subsection{Requirements for the \simupop{} implementation}

The simulations are implemented the Python package \ftprime{} 0.0.6rc0
which is available at \url{https://www.github.com/ashander/ftprime}.
Its requirements are encoded in an \texttt{environment.yml} file, compatible
with conda, but \msprime{} must be updated to the version listed below via pip
from GitHub \jda{TODO add command}.
In brief the requirements are:

\begin{enumerate}
    \item \msprime{} \jda{TODO add version from KRT's dev machine}
    \item \simupop{} 1.1.8.3
    \item python 3.5
\end{enumerate}
